Locality and Computational Reliability of Linear Response Calculations for Molecular Systems

### 2019-02-19 AnalyticStructure_main.tex
implemented proofs corrections

For a theoretical approach that is based on numerical calculations, for which no analytic reference solution exists,
reducing the computational uncertainty is the only possible way to shed light on the predictive power of the
model. In other terms, the ``accuracy'' of a result with respect to experimental data may be reliably quantified
only when the computational uncertainty is guaranteed to be significantly lower than the observed discrepancy.
For communities employing Density Functional Theory (DFT), the need of such a ``calibration'' of computer codes has led to the DeltaCodes project~\cite{deltaTest2016}.
This project represents a remarkable example of precision-driven initiative where a grassroot community of code-owners struggled to extract
well-defined computational results in the context of DFT calculations. Ground-State quantum mechanical quantities
as lattice constants and bulk modulus of elementary crystals extracted from different codes were compared with each other, with the objective of reducing the uncertainties on such data, and in some sense, to define a \emph{calibration} procedure for a Solid-State Physics DFT code.
Any other code able to extract the same quantities with the same approximations may be compared with the
results of the community in order to assess its computational reliability.

@@

In this paper, we focus our attention on Linear-Response treatments of \emph{molecular} systems,
in particular referring to LR-TDDFT calculation. We will investigate the impact that \emph{open} (i.e. Isolated) Boundary Conditions (BC), which are imposed to the system, will have on the computational evaluation of LR quantities.
We perform analytic analysis on the equations beyond LR treatment and discuss the interplay between reproducible LR quantities and the \emph{locality} of the related objects. Such considerations will enable us to distinguish quantities which \textit{a priori} can be compared among different treatments - and therefore tested for reproducibility - from others which would \emph{explicitly} depend of the employed computational treatment.

It is indeed relatively easy to have converged indicators for numerical Ground State (GS) quantities: on one side, the variational theorem
guarantees that the lower the GS energy, the more precise the numerical result; on the other side, GS quantities of molecular system  can be expressed only in terms of localized (i.e. bound) states. These considerations make relatively easy to increase the basis set completeness in view of the extraction of reference results.

We will see in what follows that the situation is not as simple in the context of LR. Indeed the response density -- even for molecular systems -- looses its localized behavior for high frequencies, independently on the nature of the specific system and perturbation. Furthermore, we will demonstrate that to increase the basis set completeness, delocalized oscillatory degrees of freedom cannot be excluded \textit{a priori}, even in the localized regime.

Lastly, we will show that the reliability of the convergence criteria that can be established for LR -- where no variational theorem exists -- depends explicitly on the target frequency range and on the features of the numerical basis set.

The paper is organized as follows.
Initially we will inspect how the question of reproducibility may be expressed
for ground state calculations of molecules.
Following these guidelines, we will then move to LR equations,
by first considering the equations of motion of the response density operator, given
the application of a perturbing field.
Then we will consider the behavior of the ``free oscillations'', i.e. the excited states
of the molecule.
Our considerations will prove to be useful for the understanding of the \emph{analytic structure} of LR quantities like
the linear susceptibility of the system (i.e. the reducible polarizability).
We support our considerations with numerical results on simple LR-TDDFT quantities.
For readability purposes, the details of each calculation are explained in Appendix~\ref{compdetails}.

Although we refer, as anticipated, to TDDFT models, such considerations might also be transposed to Many-Body Perturbation Theory
approaches, in the context of Quasi-Particle equations.

@@

Let us start our discussion by reviewing some well-known concepts of quantum mechanic under a perspective that will turn out to be very useful
in the forthcoming discussions about linear-response quantities. In particular we revisit the impact that the Isolated BC have on the solutions
of the Schr\"odinger Equation.

Consider a (one-body) wavefunction $\ket{\psi}$ solution of the equation
$\op H\ket{\psi}=\eps\ket{\psi}$, where we split the Hamiltonian operator into the sum of
a kinetic $\op T$ and a potential $\op V \equiv \op H - \op T$ term.
In view of discussing the impact of the BC, it is instructive to transform
the Schr\"odinger equation in a scattering problem, by writing
\be\lb{SEHelmholtz1}
\ket{\psi} = \GH(\eps)\ket{\op V\psi}\;,
\ee
where we have introduced the Green function of the Helmholtz operator $\GH(\eps) = (\eps-\op T)^{-1}$,
resolvent of the free-particle Hamiltonian $H_s \equiv \op T$.

When describing the ground state of a molecule, we suppose that the potential operator, even if non-local,
can be effectively restricted to wavefunctions that can be considered as nonzero only on a
bounded domain. In other terms, a molecule can be associated to a ``scatterer'' that has a \emph{finite}
spatial extension; if a state $\ket{\psi}$ is localized, it will be the same for the ket $\ket{\op V \psi}$, though
perhaps in a larger region.
For a computational analysis of the problem, this property is fundamental to the choice of the numerical treatment as we know that
computational basis sets
which are tailored to express asymptotically vanishing wavefunctions are better suited for the solution of the problem.

@@

\section{Fluctuation states. A localization argument for the response density}
\label{FluctuationState}
It is evident that the arguments presented above have a limited interest for the study
of Ground State quantities of molecules. By definition, all these quantities can be expressed as a
functional which \emph{entirely} depends on bound states. As an illustration it is sufficient to recall
the Hohenberg-Kohn theorem that states that the Ground-State (GS) energy is a functional of the GS
charge density $\dmnot$. For molecular systems which are electronically stable (i.e. no anionic or metastable configurations),
such a quantity is localized, in the sense defined in the above section. Stated otherwise,
\emph{there is no need to express efficiently any portion of the continuum spectrum} to have a reliable GS treatment of molecules.

Let us now inspect the case of LR calculations.
In this framework we consider the unperturbed Hamiltonian $\hnot$ and assumes that
the GS density $\dmnot$ is accessible and expressed in a given computational treatment.
The linear response formalism allows us to evaluate the modification of the expectation value of a generic observable induced by a
time (or frequency)-dependent perturbing field $\op\Phi(\omega)$ acting on the system.
This quantity, written in the frequency domain, can be expressed through the evaluation of the \emph{linear response functional}
\be\lb{LinearResponseFunctDef1}
\braket{\delta\op O}_\Phi(\omega) = \trace{\dm'_\Phi(\omega)\op O} \;,
\ee
where $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} under inspection
and we have introduced the \emph{response density operator} $\dm'_\Phi(\omega)$ that codifies the modification of $\dmnot$ induced by the perturbation $\op\Phi(\omega)$.

@@

We can therefore introduce an explicit representation of the
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} (FS)
$\ket{f_p^\Phi(\omega)} = \dm'_\Phi(\omega) \ket{\psi_p}$, that have values in the unoccupied subspace of $\hnot$.
However, non-hermiticity of $\dm'_\Phi(\omega)$ implies that the
fluctuation states are complex quantities.

An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states  are
written as a modified Sternheimer equation~\cite{mahan1980,giustino2012,giustino2014}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'_\Phi](\omega))\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains,
apart from the perturbing field, a further term $\op V'[\dm'_\Phi]$ due to the density-dependence of $\hnot$.
It is easy to see that the solution of equation \eqref{fluctuationStateEqMotion1} can be recast in a scattering problem:
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \GH(\omega+\epsilon_p)\left(\op V\ket{f_p^\Phi(\omega)} + \ket{s_p^\Phi(\omega)} \right)\;,
\ee
with the state
$
\ket{s_p^\Phi(\omega)} = \op Q_0(\op \Phi(\omega) +\op V'[\dm'_\Phi](\omega) )\ket{\psi_p}
$ as a source term.

From equation \eqref{rhoPrimeFluctuationStateDef2} we see that for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level that governs the asymptotic behavior of the associated Helmholtz kernel.
As a ulterior \emph{stringent} hypothesis, let us now assume that the source term $\ket{s_p^\Phi(\omega)}$ is spatially localized, \textit{i.e.} its evaluation can be restricted to a bounded real-space domain.
Even when such a locality argument holds true, for $\omega > |\eps_p|$,
the Helmholtz kernel behaves as a spherical wave and  gives raise to a $\ket{f^\Phi_p(\omega)}$ which is \emph{delocalized} over the entire real-space domain.
Such delocalization is an intrinsic consequence of the value of $\omega$, and does not depend on the particular choice of
$\op \Phi$.
Only for $\omega < |\eps_p|$ the kernel has an exponential damping factor and $\ket{f^\Phi_p(\omega)}$ exhibits a bound-state behavior in the long range,
once again provided that the above locality arguments on $\ket{s_p^\Phi(\omega)}$ -- hence on $\op V'$ and $\Phi$ -- are valid.

@@

\subsection{Localized behavior for static perturbations}
The locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain.
Any local operator, \emph{e.g.} a static electric field $\mathbf F$, with associated scalar potential $\Phi_s[\mathbf F](\r) = -\mathbf F \cdot \r$, would satisfy such a property.
We consider the perturbed Hamiltonian $\hnot[\dm_{\Phi_s}] + \op\Phi_s[\mathbf F]$, where $\dm_{\Phi_s}$ represents the GS density of the perturbed system, and assume that the
perturbation strength ($|\mathbf F|$ in the above example) is sufficiently small to be consistent with a LR treatment of the problem.
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed Hamiltonian can be extracted with traditional GS techniques,
since from the discussion around Eq.~\eqref{SEHelmholtz1}, they exhibit bound-state behavior.
\begin{figure}[t]
\includegraphics[scale=0.68]{Fig1_CO_statPolvsBox.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the static polarizability $\alpha_{zz}$ (blue continuous line) and of the ground state dipole $d_z$ (red dotted line)
of a $CO$ molecule as a function of the size of the computational domain. The inset evidences the relative convergence (in percent) of the two quantities w.r.t. the corresponding value associated with the largest
box.}
\end{figure}

From the linearized Schr\"odinger Equation of the perturbed problem, it is easy to see that the FS at $\omega=0$ can be expressed
in terms of the corresponding bound state of the perturbed Hamiltonian, i.e.
$\ket{f_p^{\Phi_s}(0)} \simeq \op Q_0 \ket{\psi_p^{\Phi_s}}$.
For any value of $p$, the locality of the $\ket{f_p^{\Phi_s}(0)}$ therefore directly stems from the
bound-state behavior of $\ket{\psi_p^{\Phi_s}}$. As $\omega=0$ is by hypothesis
lower than any of the $|\eps_p|$, this is consistent with the above considerations.

Thanks to such observations one can express quantities like the static polarizability tensor $\alpha_{ij}$ (see e.g.\cite{DebElecField}) from GS calculations of the perturbed Hamiltonian.
By expressing the induced dipole $\braket{\delta \op{\mathbf r}}_{\Phi_s}$ in the form Eq.~\eqref{LinearResponseFunctDef1}, and identifying the static response density as $\dm'_{\Phi_s}(0) = \dm_{\Phi_s} -\dmnot $, we have
\be \label{staticalpha}
\alpha_{ij} =
-\frac{1}{F_j} \left(\trace{\op r_i \dm_{\Phi_s[F_j]}} - d^{\text{(GS)}}_i \right)\;,
\ee
where $\mathbf d^\text{(GS)}=\braket{\mathbf r}$ is the ground state dipole.
Fig.~\eqref{co_alphaStatic} illustrates this procedure for a $CO$ molecule
perturbed by a static electric field.
Care has been taken in keeping the field strengths $F_j$ small enough to
preserve the validity of the linear response regime.
The local character of the fluctuation states at $\omega=0$ is
probed by verifying the convergence of the statical polarizability versus
the size of the computational domain used to express the fluctuation states. The convergence rate is compared with the
one of the static dipole of the molecule computed at zero external field.
The results show that the typical sizes of $\ket{f^\Phi_p(0)}$ are analogous to those of the unperturbed KS orbitals, which is a direct proof of the bound-state behavior of the FS.

@@

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$
exhibit bound-state long-range behavior \emph{only} when $\omega < |\eps_p|$,
while they behave as unbound states for values of $\omega$ above these threshold levels.
This fact has interesting implications in the evaluation of LR quantities. Indeed, writing the response
density in terms of FS implies that
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2 \sum_p\bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}  \;,
\ee
here we have assumed that $\op O$ is a real symmetric operator and chosen a real set of occupied molecular states.
Let us suppose that the observable of interest $\op O$ also can be restricted to localized domains in the same way as the potentials $ \op V$ and $\op \Phi$.
Once again, it would be enough to consider a local operator for this condition to be met.

Two distinct regimes can be identified, on the basis of the value of $\omega$.
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the FS are below their threshold level and equation \eqref{LinearResponseFunctDef2}
is expressed in terms of genuinely localized
quantities. At these frequencies, the linear response functional $\braket{\delta\op O}_\Phi(\omega)$
may thus be evaluated with computational setups that
are similar to those usually employed for GS calculations.
The convergence in this regime is ``strong'', in the sense that the basis functions employed enable us to express
with arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.
This is of course true even when the above states are only implicitly expressed by the employed LR treatment, as e.g. in the case of Krylov spaces generated from the states $\ket{\op\Phi(\omega) \psi_p}$ \cite{baroni2006,baroni2008,linlinKPM}.
In this regime, a computational treatment based on localized basis set is susceptible to provide a precise answer, assuming a reasonable level of completeness.

On the contrary, the computation of the linear response functional can be much more demanding in the \emph{above threshold regime},
realized for $\omega>|\eps_h|$, in which fluctuation states start behaving as unbound wave functions.
As by hypothesis $\ket{\op O \psi_p}$ is bound-state like, \emph{only} the scalar products of Eq.~\eqref{LinearResponseFunctDef2}
can be evaluated in a localized domain. In this case a computational description of fluctuation states is only meaningful in ``weak''
sense, where the bound state character of $\bra{\psi_p\op O}$ behaves as a regulator. This is naturally achieved in the above mentioned
Krylov space treatments, which explicitly deal with the expression of the scalar products. However, in this regime, imposing \emph{by design}
a localized behaviour of the FS might reveal very dangerous in view of convergence of the results since the computational basis set has
to be able to express a delocalized FS in the support domain of $\bra{\psi_p\op O}$ .
This fact implies that the degrees of completeness of the computational basis should be increased by adding
the oscillatory degrees of freedom needed to describe the FS in the relevant domain.
Stated otherwise, to achieve precise results in this regime the
underlying basis set has to be able to express delocalized states.

@@

As an example in support of these arguments we compare the {dynamical polarizability $\alpha(\omega)$} of a $CO$ molecule computed in three distinct real-space computational setups, which differ in the dimension of the computational domain.
Results reported in the two panels of figure \ref{co_spectrum} show the real and imaginary part of $\alpha(\omega)$, respectively. In all the cases a typical behavior emerges: the curves are almost coincident in the below threshold regime,
whereas the effects of the computational setup are clearly visible for values of  $\omega$ which exceed the threshold value.
For these high-frequency regions, convergence might be reached for domains of much larger sizes than those presented here~\cite{baroni2008},
which are -- as already noticed -- more than sufficient below threshold.

@@

As a consequence, we deduce that a molecular excitation \emph{may} be expressed in terms of localized states
\emph{only} when its energy is lower than \emph{all} the ionization potentials of the occupied states participating to the excitation,
namely if $\Omega_a<|\eps_p|$ $\forall p \in a$. This will only be true if the operator
$\op V'[\op E_a]$ can be restricted to localized kets and bras.
If any of these hypotheses does not hold, the excitations of the systems are defined via genuinely delocalized wavefunctions.
Furthermore, excited states undergo the orthonormalization condition
\eqref{ExcitedStateOrthNormDef1} that acts as an ulterior constraint and has the effect of determining the \emph{quantization}
of the energy levels $\Omega_a$ associated to the bound-state-like objects.
For delocalized states, as in the case of the continuum states of $\hnot$,
such a condition has to be interpreted in a distributional sense and does not lead to quantization of $\Omega_a$.
Thus we can conclude that all the states $\ket{\chi^a_p}$ and the $\ket{\phi^a_p}$ with $\Omega_a$ below the threshold value constitute
-- if localized -- a genuine \emph{discrete} set, while those in the unbound energy range behave as generalized eigenvectors associated to a
continuum spectrum.

@@

So far we have showed with both formal arguments and numerical calculations which show that the excitation spectrum splits into two sectors with very distinct features.
It is then reasonable to ask which is their relative importance and, in particular, if there is the possibility to express observable quantities, in a suitable $\omega$ range, by only employing the sector of genuinely localized excitations.
Stated otherwise, can the \emph{restriction} of $\sop \chi(\omega)$ to only localized excitations be useful to extract some physical quantities?

>>>>

For a theoretical approach that is based on numerical calculations, for which no analytic reference solution exists,
reducing the computational uncertainty is the only possible way to shed light on the predictive power of the
model. In other terms, the ``accuracy'' of a result with respect to experimental data may be reliably quantified
only when the computational uncertainty is guaranteed to be significantly lower than the observed discrepancy.
For communities employing density functional theory (DFT), the need of such a ``calibration'' of computer codes has led to the DELTACODES project~\cite{deltaTest2016}.
This project represents a remarkable example of precision-driven initiative where a grassroots community of code-owners struggled to extract
well-defined computational results in the context of DFT calculations. Ground-state quantum mechanical quantities
as lattice constants and bulk modulus of elementary crystals extracted from different codes were compared with each other, with the objective of reducing the uncertainties on such data, and in some sense, to define a \emph{calibration} procedure for a solid-state Physics DFT code.
Any other code able to extract the same quantities with the same approximations may be compared with the
results of the community in order to assess its computational reliability.

@@

In this paper, we focus our attention on linear-response (LR) treatments of \emph{molecular} systems,
in particular referring to LR-TDDFT calculation. We will investigate the impact that \emph{open} (i.e., isolated) boundary conditions (BC), which are imposed to the system, will have on the computational evaluation of LR quantities.
We perform analytic analysis on the equations beyond LR treatment and discuss the interplay between reproducible LR quantities and the \emph{locality} of the related objects. Such considerations will enable us to distinguish quantities which \textit{a priori} can be compared among different treatments - and therefore tested for reproducibility - from others which would \emph{explicitly} depend of the employed computational treatment.

It is indeed relatively easy to have converged indicators for numerical ground-state (GS) quantities: on one side, the variational theorem
guarantees that the lower the GS energy, the more precise the numerical result; on the other side, GS quantities of molecular system  can be expressed only in terms of localized (i.e. bound) states. These considerations make it relatively easy to increase the basis set completeness in view of the extraction of reference results.

We will see in what follows that the situation is not as simple in the context of LR. Indeed the response density -- even for molecular systems -- loses its localized behavior for high frequencies, independently on the nature of the specific system and perturbation. Furthermore, we will demonstrate that to increase the basis set completeness, delocalized oscillatory degrees of freedom cannot be excluded \textit{a priori}, even in the localized regime.

Lastly, we will show that the reliability of the convergence criteria that can be established for LR, where no variational theorem exists -- depends explicitly on the target frequency range and on the features of the numerical basis set.

The paper is organized as follows.
Initially, we will inspect how the question of reproducibility may be expressed
for ground state calculations of molecules.
Following these guidelines, we will then move to LR equations,
by first considering the equations of motion of the response density operator, given
the application of a perturbing field.
Then, we will consider the behavior of the ``free oscillations'', i.e. the excited states
of the molecule.
Our considerations will prove to be useful for the understanding of the \emph{analytic structure} of LR quantities like
the linear susceptibility of the system (i.e. the reducible polarizability).
We support our considerations with numerical results on simple LR-TDDFT quantities.
For readability purposes, the details of each calculation are explained in Appendix~\ref{compdetails}.

Although we refer, as anticipated, to TDDFT models, such considerations might also be transposed to many-body perturbation theory
approaches, in the context of quasiparticle equations.

@@

Let us start our discussion by reviewing some well-known concepts of quantum mechanics under a perspective that will turn out to be very useful
in the forthcoming discussions about linear-response quantities. In particular we revisit the impact that the isolated BC have on the solutions
of the Schr\"odinger equation.

Consider a (one-body) wave function $\ket{\psi}$ solution of the equation
$\op H\ket{\psi}=\eps\ket{\psi}$, where we split the Hamiltonian operator into the sum of
a kinetic $\op T$ and a potential $\op V \equiv \op H - \op T$ term.
In view of discussing the impact of the BC, it is instructive to transform
the Schr\"odinger equation in a scattering problem, by writing
\be\lb{SEHelmholtz1}
\ket{\psi} = \GH(\eps)\ket{\op V\psi}\;,
\ee
where we have introduced the Green function of the Helmholtz operator $\GH(\eps) = (\eps-\op T)^{-1}$,
resolvent of the free-particle Hamiltonian $H_s \equiv \op T$.

When describing the ground state of a molecule, we suppose that the potential operator, even if non-local,
can be effectively restricted to wave functions that can be considered as nonzero only on a
bounded domain. In other terms, a molecule can be associated to a ``scatterer'' that has a \emph{finite}
spatial extension; if a state $\ket{\psi}$ is localized, it will be the same for the ket $\ket{\op V \psi}$, though
perhaps in a larger region.
For a computational analysis of the problem, this property is fundamental to the choice of the numerical treatment as we know that
computational basis sets
which are tailored to express asymptotically vanishing wave functions are better suited for the solution of the problem.

@@

\section{Fluctuation states. A localization argument for the response density}
\label{FluctuationState}
It is evident that the arguments presented above have a limited interest for the study
of ground-state quantities of molecules. By definition, all these quantities can be expressed as a
functional which \emph{entirely} depends on bound states. As an illustration it is sufficient to recall
the Hohenberg-Kohn theorem that states that the ground-state (GS) energy is a functional of the GS
charge density $\dmnot$. For molecular systems which are electronically stable (i.e. no anionic or metastable configurations),
such a quantity is localized, in the sense defined in the above section. Stated otherwise,
\emph{there is no need to express efficiently any portion of the continuum spectrum} to have a reliable GS treatment of molecules.

Let us now inspect the case of LR calculations.
In this framework, we consider the unperturbed Hamiltonian $\hnot$ and assumes that
the GS density $\dmnot$ is accessible and expressed in a given computational treatment.
The linear response formalism allows us to evaluate the modification of the expectation value of a generic observable induced by a
time- (or frequency-) dependent perturbing field $\op\Phi(\omega)$ acting on the system.
This quantity, written in the frequency domain, can be expressed through the evaluation of the \emph{linear response functional}
\be\lb{LinearResponseFunctDef1}
\braket{\delta\op O}_\Phi(\omega) = \trace{\dm'_\Phi(\omega)\op O} \;,
\ee
where $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} under inspection
and we have introduced the \emph{response density operator} $\dm'_\Phi(\omega)$ that codifies the modification of $\dmnot$ induced by the perturbation $\op\Phi(\omega)$.

@@

We can therefore introduce an explicit representation of the
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} (FS)
$\ket{f_p^\Phi(\omega)} = \dm'_\Phi(\omega) \ket{\psi_p}$, that have values in the unoccupied subspace of $\hnot$.
However, non-Hermiticity of $\dm'_\Phi(\omega)$ implies that the
fluctuation states are complex quantities.

An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states  are
written as a modified Sternheimer equation~\cite{mahan1980,giustino2012,giustino2014}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'_\Phi](\omega))\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains,
apart from the perturbing field, a further term $\op V'[\dm'_\Phi]$ due to the density-dependence of $\hnot$.
It is easy to see that the solution of Eq. \eqref{fluctuationStateEqMotion1} can be recast in a scattering problem:
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \GH(\omega+\epsilon_p)\left(\op V\ket{f_p^\Phi(\omega)} + \ket{s_p^\Phi(\omega)} \right)\;,
\ee
with the state
$
\ket{s_p^\Phi(\omega)} = \op Q_0(\op \Phi(\omega) +\op V'[\dm'_\Phi](\omega) )\ket{\psi_p}
$ as a source term.

From equation \eqref{rhoPrimeFluctuationStateDef2} we see that for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level that governs the asymptotic behavior of the associated Helmholtz kernel.
As a ulterior \emph{stringent} hypothesis, let us now assume that the source term $\ket{s_p^\Phi(\omega)}$ is spatially localized, \textit{i.e.} its evaluation can be restricted to a bounded real-space domain.
Even when such a locality argument holds true, for $\omega > |\eps_p|$,
the Helmholtz kernel behaves as a spherical wave and  gives rise to a $\ket{f^\Phi_p(\omega)}$ which is \emph{delocalized} over the entire real-space domain.
Such delocalization is an intrinsic consequence of the value of $\omega$, and does not depend on the particular choice of
$\op \Phi$.
Only for $\omega < |\eps_p|$ the kernel has an exponential damping factor and $\ket{f^\Phi_p(\omega)}$ exhibits a bound-state behavior in the long range,
once again provided that the above locality arguments on $\ket{s_p^\Phi(\omega)}$ -- hence on $\op V'$ and $\Phi$ -- are valid.

@@

\subsection{Localized behavior for static perturbations}
The locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain.
Any local operator, \emph{e.g.} a static electric field $\mathbf F$, with associated scalar potential $\Phi_s[\mathbf F](\r) = -\mathbf F \cdot \r$, would satisfy such a property.
We consider the perturbed Hamiltonian $\hnot[\dm_{\Phi_s}] + \op\Phi_s[\mathbf F]$, where $\dm_{\Phi_s}$ represents the GS density of the perturbed system, and assume that the
perturbation strength ($|\mathbf F|$ in the above example) is sufficiently small to be consistent with a LR treatment of the problem.
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed Hamiltonian can be extracted with traditional GS techniques,
since from the discussion around Eq.~\eqref{SEHelmholtz1}, they exhibit bound-state behavior.
\begin{figure}[t]
\includegraphics[scale=0.68]{Fig1_CO_statPolvsBox.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the static polarizability $\alpha_{zz}$ (blue continuous line) and of the ground state dipole $d_z$ (red dotted line)
of a $CO$ molecule as a function of the size of the computational domain. The inset evidences the relative convergence (in percent) of the two quantities with respect to the corresponding value associated with the largest
box.}
\end{figure}

From the linearized Schr\"odinger equation of the perturbed problem, it is easy to see that the FS at $\omega=0$ can be expressed
in terms of the corresponding bound state of the perturbed Hamiltonian, i.e.
$\ket{f_p^{\Phi_s}(0)} \simeq \op Q_0 \ket{\psi_p^{\Phi_s}}$.
For any value of $p$, the locality of the $\ket{f_p^{\Phi_s}(0)}$ therefore directly stems from the
bound-state behavior of $\ket{\psi_p^{\Phi_s}}$. As $\omega=0$ is by hypothesis
lower than any of the $|\eps_p|$, this is consistent with the above considerations.

Thanks to such observations one can express quantities like the static polarizability tensor $\alpha_{ij}$ (see e.g.\cite{DebElecField}) from GS calculations of the perturbed Hamiltonian.
By expressing the induced dipole $\braket{\delta \op{\mathbf r}}_{\Phi_s}$ in the form Eq.~\eqref{LinearResponseFunctDef1}, and identifying the static response density as $\dm'_{\Phi_s}(0) = \dm_{\Phi_s} -\dmnot $, we have
\be \label{staticalpha}
\alpha_{ij} =
-\frac{1}{F_j} \left(\trace{\op r_i \dm_{\Phi_s[F_j]}} - d^{\text{(GS)}}_i \right)\;,
\ee
where $\mathbf d^\text{(GS)}=\braket{\mathbf r}$ is the ground state dipole.
Figure~\ref{co_alphaStatic} illustrates this procedure for a $CO$ molecule
perturbed by a static electric field.
Care has been taken in keeping the field strengths $F_j$ small enough to
preserve the validity of the linear response regime.
The local character of the fluctuation states at $\omega=0$ is
probed by verifying the convergence of the statical polarizability versus
the size of the computational domain used to express the fluctuation states. The convergence rate is compared with the
one of the static dipole of the molecule computed at zero external field.
The results show that the typical sizes of $\ket{f^\Phi_p(0)}$ are analogous to those of the unperturbed KS orbitals, which is a direct proof of the bound-state behavior of the FS.

@@

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$
exhibit bound-state long-range behavior \emph{only} when $\omega < |\eps_p|$,
while they behave as unbound states for values of $\omega$ above these threshold levels.
This fact has interesting implications in the evaluation of LR quantities. Indeed, writing the response
density in terms of FS implies that
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2 \sum_p\bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}  \;,
\ee
where we have assumed that $\op O$ is a real symmetric operator and chosen a real set of occupied molecular states.
Let us suppose that the observable of interest $\op O$ also can be restricted to localized domains in the same way as the potentials $ \op V$ and $\op \Phi$.
Once again, it would be enough to consider a local operator for this condition to be met.

Two distinct regimes can be identified, on the basis of the value of $\omega$.
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the FS are below their threshold level and Eq. \eqref{LinearResponseFunctDef2}
is expressed in terms of genuinely localized
quantities. At these frequencies, the linear response functional $\braket{\delta\op O}_\Phi(\omega)$
may thus be evaluated with computational setups that
are similar to those usually employed for GS calculations.
The convergence in this regime is ``strong'', in the sense that the basis functions employed enable us to express
with arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.
This is, of course, true even when the above states are only implicitly expressed by the employed LR treatment, as e.g. in the case of Krylov spaces generated from the states $\ket{\op\Phi(\omega) \psi_p}$ \cite{baroni2006,baroni2008,linlinKPM}.
In this regime, a computational treatment based on localized basis set is susceptible to provide a precise answer, assuming a reasonable level of completeness.

On the contrary, the computation of the linear response functional can be much more demanding in the \emph{above threshold regime},
realized for $\omega>|\eps_h|$, in which fluctuation states start behaving as unbound wave functions.
As by hypothesis $\ket{\op O \psi_p}$ is bound-state like, \emph{only} the scalar products of Eq.~\eqref{LinearResponseFunctDef2}
can be evaluated in a localized domain. In this case a computational description of fluctuation states is only meaningful in the ``weak''
sense, where the bound state character of $\bra{\psi_p\op O}$ behaves as a regulator. This is naturally achieved in the above mentioned
Krylov space treatments, which explicitly deal with the expression of the scalar products. However, in this regime, imposing \emph{by design}
a localized behavior of the FS might reveal very dangerous in view of convergence of the results since the computational basis set has
to be able to express a delocalized FS in the support domain of $\bra{\psi_p\op O}$ .
This fact implies that the degrees of completeness of the computational basis should be increased by adding
the oscillatory degrees of freedom needed to describe the FS in the relevant domain.
Stated otherwise, to achieve precise results in this regime the
underlying basis set has to be able to express delocalized states.

@@

As an example in support of these arguments we compare the {dynamical polarizability $\alpha(\omega)$} of a $CO$ molecule computed in three distinct real-space computational setups, which differ in the dimension of the computational domain.
Results reported in the two panels of Fig. \ref{co_spectrum} show the real and imaginary parts of $\alpha(\omega)$, respectively. In all the cases, a typical behavior emerges: the curves are almost coincident in the below threshold regime,
whereas the effects of the computational setup are clearly visible for values of  $\omega$ which exceed the threshold value.
For these high-frequency regions, convergence might be reached for domains of much larger sizes than those presented here~\cite{baroni2008},
which are, as already noticed, more than sufficient below threshold.

@@

As a consequence, we deduce that a molecular excitation \emph{may} be expressed in terms of localized states
\emph{only} when its energy is lower than \emph{all} the ionization potentials of the occupied states participating to the excitation,
namely, if $\Omega_a<|\eps_p|$ $\forall p \in a$. This will only be true if the operator
$\op V'[\op E_a]$ can be restricted to localized kets and bras.
If any of these hypotheses does not hold, the excitations of the systems are defined via genuinely delocalized wave functions.
Furthermore, excited states undergo the orthonormalization condition
\eqref{ExcitedStateOrthNormDef1} that acts as an ulterior constraint and has the effect of determining the \emph{quantization}
of the energy levels $\Omega_a$ associated to the bound-state-like objects.
For delocalized states, as in the case of the continuum states of $\hnot$,
such a condition has to be interpreted in a distributional sense and does not lead to quantization of $\Omega_a$.
Thus we can conclude that all the states $\ket{\chi^a_p}$ and the $\ket{\phi^a_p}$ with $\Omega_a$ below the threshold value constitute
-- if localized -- a genuine \emph{discrete} set, while those in the unbound energy range behave as generalized eigenvectors associated to a
continuum spectrum.

@@

So far, we have showed with both formal arguments and numerical calculations which show that the excitation spectrum splits into two sectors with very distinct features.
It is then reasonable to ask which is their relative importance and, in particular, if there is the possibility to express observable quantities, in a suitable $\omega$ range, by only employing the sector of genuinely localized excitations.
Stated otherwise, can the \emph{restriction} of $\sop \chi(\omega)$ to only localized excitations be useful to extract some physical quantities?
